---
title: "Beautiful R Tables"
author: "Jason Gavrilis"
editor: source
execute: 
  echo: false
  output: true
  warning: false
  cache: true
prefer-html: true
format:
  html:
    theme: litera
    toc: true
    toc-location: left
    number-sections: false
    number-depth: 2
    code-tools: true
    code-fold: show
    fig-dpi: 600
    fig-align: center
    fig-height: 7
    fig-width: 9
    highlight-style: espresso
    embed-resources: true
bibliography: references.bib
---

```{r warning=FALSE, echo=FALSE}
#load packages
library(pacman)
pacman::p_load(tidyverse,flextable, officer, webshot)
```

# Tables in R

## Base R

-   Base R[^1] doesn't do a good job creating presentation ready tables.

-   For example, lets look at a basic table of the top 10 fastest production cars by acceleration.

    ------------------------------------------------------------------------

[^1]: The [R Base](https://rdrr.io/r/base/base-package.html) package contains the basic functions which let **R** function as a language

```{r echo=FALSE}
cars <- read.csv("fastest_cars.csv",
                 check.names=FALSE) %>%
  as_tibble() %>% 
  filter(Rank <= 10)
cars
```

------------------------------------------------------------------------

-   What if you wanted to put this table in a presentation or report?

    -   you could export the data as text (e.g. csv) and create a table manually?

    -   you could take a screenshot and paste it into a document?

    -   you could copy each value across to another document individually?

-   None of these are good options 😥. Especially if the data changes and you have to do the whole process again! 😠.

-   Luckily, many R packages can help to create tables.

-   For a package that allows beautiful tables that are highly compatibility with Microsoft Word/PowerPoint, **Flextable** is one of the best.

## Flextable

-   [Flextable](Using%20the%20flextable%20R%20package%20(ardata-fr.github.io)) is an R package developed by *David Gohel* that *"creates pretty tables for 'HTML', 'PDF', 'Microsoft Word', 'Microsoft PowerPoint."*
-   Read on to see some examples[^2].

[^2]: Inspiration for the sophisticated example was taken from [@ardata]

### Simple Example

Let's start by passing the Sound Processor data into the flextable function.

```{r echo=TRUE}
#pass data to flextable function
ft <- flextable(ci_processors)
ft

```

Ok. Not a a bad start at all! Definitely an improvement and arguably good enough to include in a report.\
\
But what if we make a few minor improvements.

```{r echo=TRUE}
#touch up table with alignment, cell formatting, and footnotes / captions
ft <-
  flextable(ci_processors) %>% 
  theme_vanilla() %>% 
  align(align = "center", part = "all") %>%  
  align(align = "left", j = "Sound Processor", part = "all") %>% 
  colformat_num(
    big.mark = "", decimal.mark = ",",
    na_str = "na") %>% 
  autofit(add_w = .1, unit = "cm") %>% 
  footnote(., i = 1 , j = 3,
               value = as_paragraph("Year released in Australian market"),
               ref_symbols = c("1"),
               part = "header")
ft
```

Ok. This looks pretty good. And ready to publish!\
\
So how do we get this table into Word or Powerpoint?

There are a few options.

#### Exporting tables

Flextable provides three functions to easily export tables:

1.  ***save_as_image()***

    -   Enables saving to an image (e.g. .png) which can then be imported into Word or Powerpoint.

2.  ***save_as_docx()***

    -   Exports directly into a Word (docx) file.

3.  ***save_as_pptx()***

    -   Exports directly into a Powerpoint (pptx) file.

```{r echo=TRUE, eval=TRUE, output = FALSE}
# save as image zoom (defaults to a scale level of 2) 
# and expand (adds whitespace pixels around the 
# cropped table image, and has a default value of 5)
save_as_image(ft, path = "output/ci_processor_table_simple.png", 
              zoom = 4, 
              expand = 5)

#save directly to word
save_as_docx(
  `Sound Processors` = ft,
  path = "output/ci_processor_table_simple.docx", 
  pr_section = prop_section(
  page_size = page_size(
    orient = "portrait",
    width = 12, height = 8
  ),
  type = "continuous",
  page_margins = page_mar()
)
)

save_as_pptx(`Sound Processors` = ft, 
               path = "output/ci_processor_table_simple.pptx")
```

### Sophisticated Example

First we start by adding to the Sound Processor data .

```{r echo=FALSE}
#create data table adding columns with images
ci_processors <-
  tibble(
    `Sound Processor` = c(
      "Freedom", "Nucleus 5", "Nucleus 6", "Kanso", "Nucleus 7", "Kanso 2", "Nucleus 8" 
    ),
    `Released` = c(2005, 2009, 2013, 2017,2017, 2020, 2022),
    `Automatic sound Processing` = c(
      "NA", 
      "NA", 
      "SmartSound iQ with SCAN",
      "SmartSound iQ with SCAN",
      "SmartSound iQ with SCAN",
      "SmartSound iQ with SCAN",
      "SmartSound iQ 2 with SCAN 2"),
    pic = c(
      "freedom.png", "n5.png",
      "n6.png","kanso.png",
      "n7.png","kanso2.png","n8.png"
    ),
    logo = c("c_logo.png", "c_logo.png", 
             "c_logo.png", "c_logo.png", 
             "c_logo.png", "c_logo.png",
             "c_logo.png"
   )
   )%>%
  mutate(
    logo = file.path("logo", logo),
    pic = file.path("pic", pic)
  )
ci_processors
```

Then we convert the data to a flextable.

```{r echo=TRUE}
#format table output
ft <- flextable(ci_processors,
  col_keys = c("logo",
               "Sound Processor",
               "Released", 
               "Automatic sound Processing"
  )
) %>% 
  colformat_num(
    big.mark = "", decimal.mark = ",",
    na_str = "na")
ft
```

Next we add the logo image into the table.

```{r echo=TRUE}
#add logo image
ft <- compose(ft, 
    j = "logo",
    value = as_paragraph(
      as_image(src = logo, 
               width = 0.5, 
               height = 0.5)
    )
  ) %>%  
  set_header_labels(logo = "Company")

ft
```

Then we add the sound processor images into the table.

```{r echo=TRUE}
#add sound processor images
ft <- compose(ft, 
    j = "Sound Processor",
    value = as_paragraph(
      as_image(src = pic, width = .33, 
               height = 0.33, 
               guess_size = TRUE),
      " ", 
      as_chunk(x = `Sound Processor`)
    )
  )
ft
```

We then add some final touches!

```{r echo=TRUE}
#add final touches
ft <- theme_vanilla(ft) %>% 
  align(align = "center", part = "all") %>%  
  align(align = "left", j = "Sound Processor", part = "all") %>%  
  autofit(add_w = .1, unit = "cm")

ft %>% 
   footnote(., i = 1 , j = 3,
               value = as_paragraph("Year released in Australian market"),
               ref_symbols = c("1"),
               part = "header")
```

And we have a beautiful table ready to show off! 🥳👏\

```{r eval=TRUE, output = FALSE}
# save as image zoom (defaults to a scale level of 2) 
# and expand (adds whitespace pixels around the 
# cropped table image, and has a default value of 5)
save_as_image(ft, path = "output/ci_processor_table.png", 
              zoom = 4, 
              expand = 5)

#save directly to word
save_as_docx(
  `Sound Processors` = ft,
  path = "output/ci_processor_table.docx", 
  pr_section = prop_section(
  page_size = page_size(
    orient = "portrait",
    width = 12, height = 8
  ),
  type = "continuous",
  page_margins = page_mar()
)
)

#this exports but unfortunately images are not displayed.
#this is limitation of powerpoint. 
save_as_pptx(`Sound Processors` = ft, 
               path = "output/ci_processor_table.pptx")
```

# Resources

#### Flextable:

-   R documentation located at: [flextable package - RDocumentation](https://www.rdocumentation.org/packages/flextable/versions/0.7.0)

-   The help pages are in a bookdown located at: <https://ardata-fr.github.io/flextable-book/>

-   A gallery of examples is available here: <https://ardata.fr/en/flextable-gallery/>

#### Other table options:

-   [How to Make Beautiful Tables in R (rfortherestofus.com)](https://rfortherestofus.com/2019/11/how-to-make-beautiful-tables-in-r/)

[@base][@flextable][@ardata]
